---
title: "Lanczos methods for penalized and marginal likelihood models"
author: "James T. Thorson"
output: rmarkdown::html_vignette
#bibliography: 'CSPDM_vignettes.bib'
vignette: >
  %\VignetteIndexEntry{Lanczos for penalized likelihood models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
vignette_start = Sys.time()
```

This vignette demonstrates how Lanczos methods can be used to approximate standard errors and epsilon bias-correction for penalized likelihood models, without directly forming or inverting the Hessian matrix representing associations among random effects. 

```{r setup, warning = FALSE}
library(Matrix)
library(RTMB)
library(lanczosRTMB)
library(numDeriv)
```

# Generalized linear mixed model

We first demonstrate Lanczos methods using a simple generalized linear mixed model.

## Simulate from a GLMM

We first simulate data from a compound lognormal-Gamma process:

```{r settings}
set.seed(123)
n = 30
n_sum = 3

# log-linked normal deviates
u = 0 + rnorm(n)

# Conditional gamma process
y = rgamma( n, shape = 1/0.5^2, scale = exp(u) * 0.5^2 )
```

We then define a function to compute the joint likelihood, while also allowing us to calculate other outputs for later use:

```{r}
nll = function(p){
  sumexpu = sum(exp(p$u[seq_len(n_sum)]))
  ADREPORT( sumexpu )
  REPORT( sumexpu )
  nll1 = dnorm(p$u, mean=p$mu, sd=exp(p$logsd), log=TRUE)
  nll2 = dgamma( 
    y, 
    shape = 1/exp(2*p$logcv), 
    scale = exp(p$u) * exp(2*p$logcv), 
    log=TRUE
  )
  jnll = -1 * ( sum(nll1) + sum(nll2) )
  if(what == "jnll") return(jnll)
  if(what == "sumexpu") return(sumexpu)
  if(what == "biascorr") return(jnll + p$eps * sumexpu)
}

# Starting list of parameters
parlist = list(u=u*0, mu = 0, logsd = 0, logcv = 0, eps = 0)
```

Finally, we fit this model as a log-linked generalized linear mixed model (GLMM): 

```{r}
# Build RTMB object
what = "jnll"
obj = RTMB::MakeADFun( 
  nll, 
  parlist,
  map = list(eps = factor(NA)),
  random = "u",
  silent = TRUE
)

# optimize 
opt = nlminb( obj$par, obj$fn, obj$gr )
opt
```
## Fit as marginal likelihood without Cholesky decomposition

We first show that we can re-fit the model using the marginal likelihood, evaluated using Hutchinson-Lanczos instead of the Cholesky factorization that is default in TMB:

```{r}
parlist$eps = numeric()
pen0 = lanczos_MakeADFun(
  nll, 
  parlist,
  random = "u",
  k = 10,
  silent = TRUE
)

# optimize 
opt_pen0 = nlminb( pen0$par, pen0$fn )
opt_pen0
```

## Fit as a penalized likelihood model

Next, we show that the model can be refitted using penalized likelihood, conditional upon fixed values for variance parameters:

```{r}
# Define RTMB object for penalized likelihood
newmap = list(
  mu = factor(NA), 
  logsd = factor(NA), 
  logcv = factor(NA),
  eps = factor(NA)
)
pen = RTMB::MakeADFun( 
  nll, 
  obj$env$parList(), 
  map = newmap,
  silent = TRUE
)

# Re-optimize
opt_pen = nlminb( pen$par, pen$fn, pen$gr )
```

Alternatively, we provide a Newton optimizer using a truncated conjugate gradient, which operates using Hessian-vector products without ever forming the full Hessian matrix:

```{r}
# Extract tape and make Hessian-vector-product function
tape = GetTape(pen)
Hq = make_Hq( tape, pen$par )

# run Newton optimizer
opt_pen2 = newton_CG(
  par = pen$par,
  fn = tape,
  gr = tape$jacfun(),
  Hq = Hq
)
```

We can then use stochastic trace estimation and the Lanczos method to approximate the log-determinant of the Hessian for random effects:

```{r}
# log-determinant using Lanczos
Hq = make_Hq( GetTape(pen), opt_pen$par )
lanczos_logdet( Hq, k = 10, m = 3 )

# log-determinant for marginal likelihood
H = obj$env$spHess(par = obj$env$last.par.best, random = TRUE)
Matrix::determinant( H )$modulus
```
Using this log-determinant, we can also compare the log-marginal likelihood using Lanczos with the full calculation:

```{r}
# Marginal likelihoods using Lanczos
lanczos_nll( pen, k = 10, m = 10 )

# Marginal likelihood using Laplace
opt$obj
```

## Delta methods

Alternatively, we can apply Lanczos methods to standard errors for parameters.  Here, we will sample the first three random effects

```{r}
samples = lanczos_sample(
   Hq = Hq,
   q = c( rep(1,3), rep(0,n-3) ),
   k = 30,
   n = 1000,
   orthogonalize = TRUE
)

#
sdrep = sdreport( obj, ignore.parm.uncertainty = TRUE )
as.list(sdrep, what = "Std. Error")$u[1:3]
apply( samples, MARGIN = 1, FUN = sd)[1:3]
```
Alternatively, we can apply Lanczos methods to approximate the Hessian with respect to a derived quantity.  This involves calculating the gradient for the derived quantity with respect to coefficients:

```{r}
what = "sumexpu"
grad = RTMB::MakeADFun( 
  nll, 
  obj$env$parList(), 
  map = newmap,
  silent = TRUE
)$gr( pen$par )[1,]
```

We can then use this gradient to approximate samples from the Hessian matrix and use that to calculate a Monte Carlo estimator for standard errors:

```{r}
samples = lanczos_sample(
   Hq = Hq,
   q = grad,
   k = 30,
   n = 1000,
   orthogonalize = TRUE
)

# Compare SD of samples with the delta method
sumexpu1 = apply( samples, MARGIN = 2, FUN = \(v)pen$report(v)$sumexpu )
c( pen$report()$sumexpu, sd(sumexpu1) )
summary(sdrep)['sumexpu',]
```

Alternatively, we can calculate the standard error for a derived quantity using Lanczos within the delta method:

```{r}
# 
Var = lanczos_variance(
  Hq = Hq,
  q = grad,
  k = 3
)

# Compare with the delta method
c( pen$report()$sumexpu, sqrt(Var) )
summary(sdrep)['sumexpu',]
```

## Epsilon methods

Finally, we can use the gradient of the log-likelihood with respect to a dummy variable epsilon to correct for retransformation bias:

```{r}
# One-sided finite difference
what = "biascorr"
phat = obj$env$parList()
phat$eps = 0.0001
pen_hi = RTMB::MakeADFun( 
  nll, 
  parameters = phat, 
  map = newmap,
  silent = TRUE
)

# Re-optimize
opt_hi = optim( 
  pen_hi$par, pen_hi$fn, pen_hi$gr, 
  method = "L-BFGS-B", control = list(factr = 1e-2) 
)

# Calculate marginal nll for finite-difference
nll_hi = lanczos_nll( pen_hi, k = 10, m = 10, seed = 123 )
nll_mid = lanczos_nll( pen, k = 10, m = 10, seed = 123 )

# Epsilon bias-correction estimator
sdrep = sdreport( obj, bias.correct = TRUE )

#
(nll_hi['nll'] - nll_mid['nll']) / (phat$eps)
summary(sdrep)['sumexpu',]
```

# Spatial change of support

Next, we fit a model that involves spatial change of support, which breaks sparsity in the inner Hessian.  We again simulate data from the change-in-support process:

```{r, include = TRUE, warning=FALSE, message=FALSE}
# Settings
set.seed(123)
nx = 10^1
ny = 10^2
rho = 0.9

# Simulate AR1 process approaching random walk (i.e., ill-conditioned inner problem)
Px = bandSparse( n = nx, k = c(-1,1), diagonals = list(rep(0.5,nx),rep(0.5,nx)) )
Py = bandSparse( n = ny, k = c(-1,1), diagonals = list(rep(0.5,ny),rep(0.5,ny)) )
Ix = Diagonal(n=nx)
Iy = Diagonal(n=ny)
Qx = (Ix - rho * t(Px)) %*% (10 * Ix) %*% (Ix - rho * Px)
Qy = (Iy - rho * t(Py)) %*% (10 * Iy) %*% (Iy - rho * Py)
Q = kronecker( Qy, Qx )
x = RTMB:::rgmrf0( n= 1, Q = Q )[,1]
y = rpois( nx*ny, lambda = exp(1 + x) )
which_seen = sample( seq_len(nx*ny), size = nx*ny/100, replace = FALSE)
sumy = sum(y)
#y[-which_seen] = NA
```


We then define the joint likelihood that includes a term for the total across samples:
```{r, include = TRUE, warning=FALSE, message=FALSE}
nll = function(p){
  Qx = (Ix - plogis(p$invlogis_rho) * t(Px)) %*% (exp(2*p$logtau) * Ix) %*% (Ix - plogis(p$invlogis_rho) * Px)
  Qy = (Iy - plogis(p$invlogis_rho) * t(Py)) %*% (exp(2*p$logtau) * Iy) %*% (Iy - plogis(p$invlogis_rho) * Py)
  Q = kronecker( Qy, Qx )
  loglik1 = dgmrf(p$x, Q = Q, log = TRUE)
  loglik2 = sum(dpois(y, exp(p$mu + p$x), log=TRUE), na.rm=TRUE)
  loglik3 = dnorm(log(sumy), log(sum(exp(p$mu + p$x))), sd = 0.01, log = TRUE)
  -1 * ( loglik1 + loglik2 + loglik3 )
}
parlist = list( x=rnorm(nx*ny), logtau = 0, invlogis_rho = 0, mu = 0 )
```

We then fit this as normal in RTMB:

```{r, include = TRUE, warning=FALSE, message=FALSE}
start_time = Sys.time()
obj_RTMB = MakeADFun(
  nll,
  parlist,
  random = "x",
  silent = TRUE
)
opt_RTMB = nlminb( obj_RTMB$par, obj_RTMB$fn, control = list(trace = 1) )
opt_RTMB$run_time = Sys.time() - start_time
```

We can see that the Hessian is dense:
```{r, include = TRUE, warning=FALSE, message=FALSE}
Matrix::image( obj_RTMB$env$spHess(random=TRUE) )
```

Alternatively, we can fit this using Lanczos methods, which are less sensitive to the lack-of-sparsity.  Here, we avoid constructing the gradient of the Laplace-Lanczos approximation to the log-marginal likelihood, pending improvements:

```{r, include = TRUE, warning=FALSE, message=FALSE}
start_time = Sys.time()
obj = lanczos_MakeADFun(
  nll,
  parlist,
  random = "x",
  k = 40,
  make_gr = FALSE,
  silent = TRUE
)
opt = nlminb(
  obj$par,
  \(x) obj$fn(x),
  control = list(trace = 1)
)
opt$run_time = Sys.time() - start_time
```

Where we can compare the runtime for the two models:
```{r, eval=TRUE, echo=TRUE, message=FALSE, fig.width=6, fig.height=6}
runtime = c(
  RTMB = opt_RTMB$run,
  Lanczos = opt$run
)
knitr::kable( runtime, digits=2, caption="Run-times" )
```

```{r, include = FALSE, warning=FALSE, message=FALSE}
run_time = Sys.time() - vignette_start
```
Runtime for this vignette: `r paste( round(unclass(run_time),2), attr(run_time, "units") )`

## Works cited
